---
title: "Lab 08 - Text Mining/NLP"
output: html_document
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(eval = F, include  = T)
```

# Learning goals

- Use `unnest_tokens()` and `unnest_ngrams()` to extract tokens and ngrams from text
- Use dplyr and ggplot2 to analyze and visualize text data
- Try a theme model using `topicmodels`

# Lab description

For this lab we will be working with the medical record transcriptions from https://www.mtsamples.com/ available at https://github.com/JSC370/JSC370-2025/tree/main/data/medical_transcriptions.

# Deliverables

1. Questions 1-7 answered, knit to pdf or html output uploaded to Quercus.

2. Render the Rmarkdown document using `github_document` and add it to your github site. Add link to github site in your html.


### Setup packages

You should load in `tidyverse`, (or `data.table`), `tidytext`, `wordcloud2`, `tm`, and `topicmodels`.


## Read in the Medical Transcriptions

Loading in reference transcription samples from https://www.mtsamples.com/

```{r eval=TRUE}
library(tidytext)
library(tidyverse)
library(wordcloud2)
library(tm)
library(topicmodels)

mt_samples <- read_csv("https://raw.githubusercontent.com/JSC370/JSC370-2025/main/data/medical_transcriptions/mtsamples.csv")
mt_samples <- mt_samples |>
  select(description, medical_specialty, transcription)

head(mt_samples)
```

---

## Question 1: What specialties do we have?

We can use `count()` from `dplyr` to figure out how many different medical specialties are in the data. Are these categories related? overlapping? evenly distributed? Make a bar plot.

The distribution of medical specialties in the dataset is highly uneven, with Surgery having the most transcriptions, followed by Orthopedic, Radiology, and General Medicine, while many other specialties have significantly fewer entries. Some categories may overlap, such as General Medicine potentially including cases from multiple specialties, and SOAP/Chart/Progress Notes covering diverse medical fields.

```{r eval=TRUE}
mt_samples |>
  count(medical_specialty, sort = TRUE) |>
  ggplot(aes(x = n, y = fct_reorder(medical_specialty, n))) +
  geom_col(fill = "lightblue") +
  theme_minimal()
```

---

## Question 2: Tokenize

- Tokenize the the words in the `transcription` column
- Count the number of times each token appears
- Visualize the top 20 most frequent words with a bar plot
- Create a word cloud of the top 20 most frequent words

### Explain what we see from this result. Does it makes sense? What insights (if any) do we get?
The chart shows that the most frequent words in the medical transcription dataset are common stopwords like "the," "and," "was," and "of," which do not provide meaningful insights. To extract useful medical information, we should remove stopwords and focus on domain-specific terms or phrases.
```{r eval=TRUE}
tokens <- mt_samples |>
  select(transcription) |>
  unnest_tokens(word, transcription) |>
  group_by(word) |>
  summarize(word_frequency = n()) |>
  arrange(across(word_frequency, desc)) |>
  head(20)
tokens

tokens |>
  ggplot(aes(fct_reorder(word,word_frequency),word_frequency)) +
  geom_bar(stat="identity",fill = "lightblue")+
  coord_flip()+
  theme_bw()
```

```{r}
tokens |> 
  count(word, sort=TRUE) |>
  wordcloud2(size = 0.5,color = "random-light",backgroundColor = "lightblue")
```

---

## Question 3: Stopwords

- Redo Question 2 but remove stopwords
- Check `stopwords()` library and `stop_words` in `tidytext`
- Use regex to remove numbers as well
- Try customizing your stopwords list to include 3-4 additional words that do not appear informative

### What do we see when you remove stopwords and then when you filter further? Does it give us a better idea of what the text is about?

When stopwords are removed, the most frequent words become more domain-specific, revealing terms like "pain," "surgery," "treatment," and "patient", which are more relevant to medical transcriptions. 

```{r eval=TRUE}
head(stopwords("english"))
length(stopwords("english"))
head(stop_words)



tokens2 <- mt_samples |>
  select(transcription) |>
  unnest_tokens(word, transcription) |>
  anti_join(stop_words, by = "word") |>
  filter(!str_detect(word, "^[0-9]+$")) |> 
  filter(!word %in% c("mm","mg","noted")) |>
  group_by(word) |>
  summarize(word_frequency = n()) |>
  arrange(across(word_frequency, desc)) |>
  head(20)
tokens2

tokens2 |>
  ggplot(aes(fct_reorder(word,word_frequency),word_frequency)) +
  geom_bar(stat="identity",fill = "lightblue")+
  coord_flip()+
  theme_bw()

tokens2 |> 
  count(word, sort=TRUE) |>
  wordcloud2(size = 0.4,color = "random-light",backgroundColor = "lightblue")
```


---



## Question 4: ngrams

Repeat question 2, but this time tokenize into bi-grams. How does the result change if you look at tri-grams? Note we need to remove stopwords a little differently. You don't need to recreate the wordclouds.

The bi-gram analysis provides clearer medical context, revealing key terms like "blood pressure," "medical history," and "postoperative diagnosis", which are more informative than single words. This helps identify common themes in medical documentation, and further refining with tri-grams or categorization could enhance insight extraction.

```{r eval=TRUE}
stop_words2 <- c("en", "mm", "mg", "count", stop_words$word)
sw_start <- paste0("^", paste(stop_words2, collapse=" |^"), "$")
sw_end <- paste0("", paste(stop_words2, collapse="$| "), "$")

tokens_bigram <- mt_samples |>
  select(transcription) |>
  unnest_tokens(ngram, transcription, token = "ngrams", n = 2) |>
  filter(!grepl(sw_start, ngram, ignore.case = TRUE))|>
  filter(!grepl(sw_end, ngram, ignore.case = TRUE))|>
  filter(!grepl("[[:digit:]]+", ngram))|>
  group_by(ngram) %>% 
  summarize(word_frequency = n()) %>% 
  arrange(across(word_frequency, desc)) %>% 
  head(20)

tokens_bigram %>% 
  ggplot(aes(ngram, word_frequency)) +
  geom_col(fill="dodgerblue") +
  coord_flip() +
  theme_bw()
```
---

## Question 5: Examining words

Using the results from the bigram, pick a word and count the words that appear before and after it, and create a plot of the top 20.

```{r eval=TRUE}
library(stringr)
# e.g. patient, blood, preoperative...
tokens_bigram |>
  filter(str_detect(ngram, regex("\\sblood$|^blood\\s"))) %>% 
    mutate(word = str_remove(ngram, "blood"),
         word = str_remove_all(word, " ")) |>
  group_by(word) %>% 
  head(20) %>% 
  ggplot(aes(reorder(word, word_frequency), word_frequency)) +
  geom_col(fill="dodgerblue") +
  theme_bw()
```

---


## Question 6: Words by Specialties

Which words are most used in each of the specialties? You can use `group_by()` and `top_n()` from `dplyr` to have the calculations be done within each specialty. Remember to remove stopwords. How about the 5 most used words?


```{r eval=TRUE}
mt_samples |>
   unnest_tokens(word, transcription) |>
   anti_join(stop_words, by = "word") |>
  filter(!str_detect(word, "^[0-9]+$")) |> 
  filter(!word %in% c("mm","mg","noted")) |>
   group_by(medical_specialty) |>
   count(word,sort = TRUE)|>
  top_n(5,n)
```


## Question 7: Topic Models

See if there are any themes in the data by using a topic model (LDA). 

- you first need to create a document term matrix
- then you can try the LDA function in `topicmodels`. Try different k values.
- create a facet plot of the results from the LDA (see code from lecture)


```{r eval=TRUEE}
transcripts_dtm <- mt_samples |>
  select(transcription) |>
  unnest_tokens(word, transcription) |>
  anti_join(stop_words, by = "word") |>
  filter(!str_detect(word, "^[0-9]+$")) |> #[[:digit:]]+
  filter(!word %in% c("mm", "mg", "noted")) |>
  DocumentTermMatrix()

transcripts_dtm <- as.matrix(transcripts_dtm)   

transcripts_lda1 <- LDA(transcripts_dtm, k = 5,
                       control = list(seed = 1234))

transcripts_lda2 <- LDA(transcripts_dtm, k = 3,
                       control = list(seed = 1234))

transcripts_top_terms1 <- 
  tidy(transcripts_lda1, matrix = "beta") |>
  filter(!str_detect(term, "^[0-9]+$")) |>
  group_by(topic) |>
  slice_max(beta, n = 10) |>
  ungroup() |>
  arrange(topic, -beta)

transcripts_top_terms2 <- 
  tidy(transcripts_lda2, matrix = "beta") |>
  filter(!str_detect(term, "^[0-9]+$")) |>
  group_by(topic) |>
  slice_max(beta, n = 10) |>
  ungroup() |>
  arrange(topic, -beta)

transcripts_top_terms1 |>
  mutate(term = reorder_within(term, beta, topic)) |>
  ggplot(aes(beta, term, fill = factor(topic))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~topic, scales = "free") +
  scale_y_reordered() +
  theme_bw()

transcripts_top_terms2 |>
  mutate(term = reorder_within(term, beta, topic)) |>
  ggplot(aes(beta, term, fill = factor(topic))) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~topic, scales = "free") +
  scale_y_reordered() +
  theme_bw()

```




